Predicting computational reproducibility of data analysis pipelines \\in large population studies using collaborative filtering

### 2018-08-17 paper.tex
A few fixes. Contracted to 10 pages

Computational reproducibility, the ability to recompute data analyses 
over time and 
space~\cite{peng2011reproducible}, has become a critical component of 
scientific methodology as many researchers acknowledge the existence of 
a reproducibility crisis~\cite{baker2016there}. Among other factors, 
such as data and code sharing, infrastructural characteristics 
play an important role in making experiments reproducible. For instance, in 
neuroinformatics, our primary field of interest, several studies have 
shown the effect of the 
operating system on computational 
results~\cite{gronenschild2012effects, glatard2015reproducibility}.
However, conducting such reproducibility studies at scale is cumbersome 
due to the computational and storage requirements of data analysis 
pipelines.

@@

The reproducibility of a given pipeline may vary across subjects, for 
instance due to different pipeline branches being executed depending on 
the nature, cardinality or content of the input data. As an example, in 
the public database released by the Human Connectome 
Project~\cite{van2013wu}, subjects may have one or two anatomical images of 
each modality; when two images are 
present, they are aligned together and averaged. The presence of 
artifacts, for instance due to high motion during the acquisition, may 
also trigger corrections that are not otherwise required. Some branches 
of the pipeline may also be executed only for specific acquisition 
parameters, for instance the intensity of the magnetic field. To 
capture all these variations, reproducibility evaluations need to be 
conducted on many subjects, which is unwieldy.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by works that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix. Instead, the training set has to 
respect the time constraints imposed by the file creation order during 
pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
Finally, we conclude on the best sampling method to use, and on the 
impact, limitations and generalizability of the results.

@@

The pipeline to be evaluated is represented by a matrix $U$ of size 
$N_f \times N_s$, where the $N_s$ columns represent data coming from 
different subjects, and the $N_f$ rows represent the files produced by 
the pipeline execution. While subjects are not ordered, files are, for 
instance from their last modification time in a sequential execution, 
and we assume that this order is consistent across subjects. $U_{i, j}$ 
measures the reproducibility of file $i$ produced during the processing 
of subject $j$ in two conditions, for instance two different 
operating systems. In our experiments, we restrict $U_{i,j}$ to be 
boolean, but our methods can be applied for real values as well.

Our goal is to predict the test set $\mathbb{T'}$ of missing values of 
$U$ from a training set $\mathbb{T}$ of known ones, where $\mathbb{T} 
\cap \mathbb{T'} = \emptyset$ and $\mathbb{T} \cup \mathbb{T'} = 
\{U_{i,j}\}$. In contrast with the traditional collaborative filtering 
problem, (1) we have control over the construction of the training set, 
for instance we can choose to include only files producde by specific 
subjects, or the first files produced by the processing of every 
subject, and (2) the construction of the training set is constrained by 
the 
order on the matrix rows, which is formalized as follows: 
\begin{equation}
\begin{array}{l}
\forall (i, j, k) \in \llbracket 1, N_f \rrbracket \times \llbracket 1, N_f \rrbracket \times \llbracket 1, N_s \rrbracket, \\
 \left( U_{i,k} \in \mathbb{T} \quad \mathrm{and} \quad U_{j,k} \in \mathbb{T'} \right) \Rightarrow i < j. \label{eq:time}
 \end{array}
\end{equation}

@@

A third class of methods, which is the one that we will apply to our 
problem, is based on the factorization of the utility matrix to 
estimate latent factors along which users and items are represented. 
This method is described 
in~\cite{koren2009matrix} and became famous as it contributed to winning the 
Netflix prize in 2009. To summarize, the method aims at finding $q_i$ 
and $p_u$ vectors of $\mathbb{R}^f$, where $f$ is the number of latent factors, such that:
\begin{equation*}
r_{ui} = q_i^Tp_u,
\end{equation*}
where $r_{ui}$ is the rating of item $i$ by user $u$. In practice, the optimization
involves a regularization term to avoid overfiting particular users or items. The method
finds $q_i$ and $p_u$ that minimize the following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2\right)
,
\end{equation*}
where $\lambda$ is the regularization parameter and $\mathbb{T}$ is the training set.

It is also common to include user biases $b_u$ and item biases $b_i$ in 
the optimization, defined as the average deviation of user $u$ and item 
$i$ to the global average $\mu$. The goal then becomes to find $q_i$ and $p_u$ that minimize the 
following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - \mu - b_u - b_i - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2 + b_u^2 + b_i^2\right)
\end{equation*}

Two good techniques to perform the optimization are stochastic gradient 
descent and alternating least squares (ALS). In the remainder of this 
paper, we use matrix factorization as implemented in Apache Spark's 
MLLib version 2.3.0, which uses ALS. We round predictions to the 
nearest integer to obtain binary values.

@@

\begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
                  \caption{Complete Columns}
                  \label{fig:Columns-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
                  \caption{Complete Rows}
                  \label{fig:Rows-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                 \includegraphics[width=\columnwidth]{figures/5vs7_random-real_04_training}
                \caption{Random Subjects (RS)}
                  \label{fig:Uniform-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_diagonal_04_training}
                  \caption{RFNU}
                  \label{fig:Diagonal-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-largest_04_training}
                  \caption{RFNTL}
                  \label{fig:triangular-L-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-smallest_04_training}
                  \caption{RFNTS}
                  \label{fig:triangular-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.45\columnwidth]{figures/5vs7_random-unreal_04_training}
                  \caption{Random Unreal}
                  \label{fig:Random-Unreal-Sample-Training-set}
        \end{subfigure}
        \caption{Sample training set for different methods with a training ratio of 0.4 (dark elements are in the training set).}
        \label{fig:sampling}
\end{figure}

@@

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]
 For $\alpha \leq 0.5$, we ensure that the average number of selected 
 files by subject is $\alpha N_f$ by sampling the number of 
 files selected in every subject from $U(0,N_f)$ with 
 probability $2\alpha$, and setting it to 0 otherwise. For $\alpha > 
 0.5$, this is ensured by the value used for $a$, which leads the 
 average of U(a, b) to be $\alpha N_f$.

@@

\subsubsection{Random Unreal (Figure 
~\ref{fig:Random-Unreal-Sample-Training-set})} The training set is 
sampled in a random uniform way, regardless of the file generation 
times. This method does not respect the time constraints in 
Equation~\ref{eq:time}, it will be used as a baseline for comparison 
with other methods.

@@

We generated synthetic matrices as shown in 
Figure~\ref{fig:synthetic-data}. Each matrix has 100 files and 100 
subjects of different \emph{types}. Subjects of the same type behave 
identically and all types contain the same number of subjects $\pm$ 1. 
Such a decomposition by subject type correspond to the case where 
different subjects may have data of different nature, as it is for 
instance the case in the neuroimaging 
dataset of the Human Connectome Project~\cite{van2013wu} where not all the 
subjects have the same amount of images.

 For subjects of a given type, the reproducibility matrix consists of 
 $log(n)$  blocks, where $n$ is the number of types. Blocks are 
 arranged with all the possible variation patterns: some types do not 
 vary at all, while other ones vary in every block. Such variation 
 patterns are meant to mimic the logic of data 
processing pipelines: each block of files represents the files produced 
at a given stage of the pipeline, which may or may not contain 
reproducibility errors depending on the subject type.

@@

Figure~\ref{fig:utility-matrices} shows the utility matrices
obtained for the PreFreesurfer and Freesufer pipeline. The computational
reproducibility of these pipelines varies across subjects,
but some files behave
consistently across all subjects, leading to complete black or white
lines.
%The ratio of positive elements in utility matrix of CentOS5 vs CentOS6(C5C6),  CentOS5 vs CentOS7(C5C7) and  CentOS6 vs CentOS7(C6C7) are
%$0.34$, $0.79$ and $0.79$ respectively.

@@

\section{Results}

\label{sec:results}

We conducted two experiments for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with subject and file biases. We 
compare the performance of our sampling methods to (1) a dummy 
classifier that always predicts the value in the majority class and (2) 
the Random Unreal method, used as the baseline sampling technique. All 
reported values are averages over 5 repetitions. Due to sampling 
issues, it is possible that the actual training ratio obtained with 
some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true.

@@

\subsubsection{ALS without Bias}
 Figure~\ref{fig:results-synthetic} shows accuracy results for ALS 
 without bias. By construction, the accuracy of the dummy classifier is 
 close to 0.5 for all subject types. Random Unreal performs very well 
 for all subject types, which confirms that ALS is working correctly. 
 Unsurprisingly, all the other methods 
perform better than the dummy classifier, and their accuracy decreases 
as the number of subject types increases.

However, only 3 methods are able to provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. 
Surprisingly, RFNTS does not perform well for more than 2 subject 
types, even for a training ratio of 0.9.

@@

Figure~\ref{fig:results-synthetic-als-bias} shows accuracy results for 
ALS with bias. RS, RFNU and RFNTL remain the methods that best compare 
to Random Unreal, but the achieved accuracy is much lower than without 
biases. This is presumably due to 
the fact that file biases are all close to 0.5 in the synthetic data. 
In such a situation, biases are detrimental to the accuracy of the 
collaborative filtering process and should not be included.

@@

Figure~\ref{fig:results-real-als} shows the accuracy results for ALS without bias. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets, when the training ratio is larger 
than 0.5. RS and RFNTL are underperforming compared to RFNU. 
Complete rows and complete columns do not even reach the accuracy of 
the dummy classifier. The reference method, Random Unreal, has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.

@@

Figure~\ref{fig:results-real-als-bias} shows the accuracy results for 
ALS with bias. From a training ratio of 0.5, all methods have a very 
good accuracy for all real datasets, except complete columns in 
Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, RFNTL and 
RFNTS also perform well for a training ratio lower than 0.5. Overall, 
the accuracy of the methods is much higher than without bias, due to 
the fact that the file bias is very strong in this dataset, as 
it can be seen in Figure~\ref{fig:utility-matrices}.

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{data/results/roc/roc.pdf}
\includegraphics[width=\columnwidth]{data/results/roc/roc-closeup.pdf}
\caption{Comparison of the sampling methods in the ROC space for the 6 synthetic datasets and Freesurfer. Right: entire dataset. Left: close-up on the top-left part.}
\label{fig:roc}
\end{figure*}

\begin{table}
\centering
\begin{tabular}{ccc}
& Sensitivity & Specificity \\
\hline
Complete Columns & 0.53 & 0.97\\
Complete Rows & 0.43 & 0.89\\
Random Subjects & 0.88 & 0.99\\
RFNU & 0.92 & 0.99 \\
RFNTL & 0.81 & 0.97\\
RFNTS & 0.65 & 0.93 \\
Random Unreal & 1 & 1
\end{tabular}
\caption{Average sensitivity and specificity values for the 6 synthetic datasets and Freesurfer ($\alpha=0.9$).}
\label{table:roc}
\end{table}

\subsection{Effect of the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in the ALS optimization for the RFNU method, and the synthetic dataset with 8 types. 
For training ratios greater than 0.5, the accuracy obtained 
with 3 factors is substantially higher than with 2 factors. Beyond 
3, the number of factors does not have an effect on the accuracy. 
This shows that the data is best explained by 3 factors. For training 
ratios lower than 0.5, the number of factors does not have any effect 
on the accuracy.

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, (RFNU, synthetic dataset, 8 types, 5 iterations).}
\label{fig:factors}
\end{figure}

Figure~\ref{fig:factors3d} confirms that 3 factors are indeed enough to 
separate the files of the synthetic dataset with 8 subject types in 3 
blocks, and the subjects in 8 types.

\begin{figure}[h]
\begin{subfigure}{\columnwidth}
\includegraphics[width=0.9\columnwidth]{./data/results/latent-factors/5i-3f-ALS-r1-08-userFactors.pdf}
\caption{Subject factors}
\end{subfigure}
\begin{subfigure}{\columnwidth}
\includegraphics[width=0.9\columnwidth]{./data/results/latent-factors/5i-3f-ALS-r1-08-itemFactors.pdf}
\caption{File factors}
\end{subfigure}
\caption{Subject and file factors produced by the collaborative filtering method (RFNU, synthetic dataset, 8 types, 3 factors, 5 iterations).}
\label{fig:factors3d}
\end{figure}

\subsection{Effect of the number of iterations}
Figure~\ref{fig:iterations} illustrates the effect of the number of 
iterations used in the ALS optimization, for the 
RFNU method, the synthetic dataset with 8 types, and 50 factors.

From a training ratio of 0.7, the number of iterations has a moderate 
effect on the accuracy. Our previous experiments used 5 iterations, 
which only degrades the accuracy slightly compared to 15 or 20 iterations.

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Iterations-rank50/ALS-8-types-iteration.pdf}
\caption{Effect of the number of iterations used in ALS, (RFNU, synthetic dataset, 8 types, 50 factors).}
\label{fig:iterations}
\end{figure}

@@

Figure~\ref{fig:overlap} compares the prediction errors made by RFNU, 
RFNTL and RFNTS. The matrices in this Figure represent the training set 
in black (negatives) and white (positives), and the test set in green 
(true positives), yellow (false negatives), gray (true negatives) and 
red (false positives). This representation provides insights regarding 
where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of prediction errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

\begin{figure*}
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNU_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNU: $U(0.8N_f, N_f)$}
        \label{fig:overlap-rfnu}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-L_Syn_8_ALS_09_test_data_matrix_run2.png}
        \caption{RFNTL: $T(0.85N_f, 0.85N_f, N_f)$}
        \label{fig:overlap-rfntl}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-S_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNTS: 0.3$T(0, N_f, N_f)$+0.7$N_f$}
        \label{fig:overlap-rfnts}
\end{subfigure}
\caption{Comparison between RFNU, RFNTL and RFNTS on synthetic dataset with 8 types, $\alpha=0.9$.}
\label{fig:overlap}
\end{figure*}

Figure~\ref{fig:error-locality} shows the results obtained by RFNU on 
the Freesurfer dataset for a training ratio of 0.6. The prediction 
error is localized (1) at the bottom of the matrix, which corresponds 
to the end of the execution, and (2) in the regions where file 
reproducibility varies across subjects, i.e., lines are not entirely 
black or entirely white.
\begin{figure}
\centering
\includegraphics[width=0.55\columnwidth]{figures/RFNU_FS100F_ALS_06_test_data_matrix_run1.png}
\caption{Prediction results for RFNU on the Freesurfer dataset, $\alpha=0.6$.}
\label{fig:error-locality}
\end{figure}

>>>>

Computational reproducibility, the ability to recompute data analyses 
over time and 
space~\cite{peng2011reproducible}, has become a critical component of 
scientific methodology as many researchers acknowledge the existence of 
a reproducibility crisis~\cite{baker2016there}. Infrastructural 
characteristics play an important role in making experiments 
reproducible, among other factors, such as data and code sharing. For 
instance, in 
neuroinformatics, our primary field of interest, several studies have 
shown the effect of the 
operating system on computational 
results~\cite{gronenschild2012effects, glatard2015reproducibility}.
However, conducting such reproducibility studies at scale is cumbersome 
due to the computational and storage requirements of data analysis 
pipelines.

@@

The reproducibility of a given pipeline may vary across subjects, for 
instance due to different pipeline branches being executed depending on 
the input data content. As an example, in 
the public database released by the Human Connectome 
Project~\cite{van2013wu}, subjects may have one or two anatomical 
images of each modality; when two images are present, they are
aligned together and averaged. Acquisition artifacts, for instance due to motion,
may also trigger corrections not otherwise required. Some 
branches of the pipeline may also be executed only for specific 
acquisition 
parameters. These remarks are 
consistent with recent findings that showed that variations in the 
results of a functional MRI analysis are dependent on the dataset being 
analyzed~\cite{bowring2018exploring}. To capture such variations, 
reproducibility evaluations need to be conducted on many subjects, 
which is unwieldy.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by works that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix defining the collaborative filtering problem. Instead, the training set has to 
respect time constraints imposed by the file creation order during 
pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are in 
Section~\ref{sec:results}.
Finally, we conclude on the best sampling method to use, and on the 
impact, limitations and generalizability of the results.

@@

The pipeline to be evaluated is represented by a matrix $U$ of size 
$N_f \times N_s$, where the $N_s$ columns represent data coming from 
different subjects, and the $N_f$ rows represent the files produced by 
the pipeline. While subjects are not ordered, files are, for 
instance from their last modification time in a sequential execution, 
and we assume that this order is consistent across subjects. $U_{i, j}$ 
measures the reproducibility of file $i$ produced during the processing 
of subject $j$ in two conditions, for instance two different 
operating systems. In our experiments, we restrict $U_{i,j}$ to be 
boolean, but our methods can be applied for real values as well.

Our goal is to predict the test set $\mathbb{T'}$ of missing values of 
$U$ from a training set $\mathbb{T}$ of known ones, where $\mathbb{T} 
\cap \mathbb{T'} = \emptyset$ and $\mathbb{T} \cup \mathbb{T'} = 
\{U_{i,j}\}$. In contrast with the traditional collaborative filtering 
problem, we have control over the construction of the training set, 
for instance we can choose to include only files producde by specific 
subjects, or the first files produced by the processing of every 
subject. Moreover, the construction of the training set is constrained by 
the 
order on the matrix rows, which is formalized as follows: 
\begin{equation}
\begin{array}{l}
\forall (i, j, k) \in \llbracket 1, N_f \rrbracket \times \llbracket 1, N_f \rrbracket \times \llbracket 1, N_s \rrbracket, \\
 \left( U_{i,k} \in \mathbb{T} \quad \mathrm{and} \quad U_{j,k} \in \mathbb{T'} \right) \Rightarrow i < j. \label{eq:time}
 \end{array}
\end{equation}

@@

A third class of methods, which is the one that we will use, is based 
on the factorization of the utility matrix to estimate latent factors 
along which users and items are represented. This method is described 
in~\cite{koren2009matrix} and became famous as it contributed to 
winning the 
Netflix prize in 2009. To summarize, the method aims at finding $q_i$ 
and $p_u$ vectors of $\mathbb{R}^f$, where $f$ is the number of latent factors, such that:
\begin{equation*}
r_{ui} = q_i^Tp_u,
\end{equation*}
where $r_{ui}$ is the rating of item $i$ by user $u$. In practice, the optimization
involves a regularization term to avoid overfiting particular users or items. The method
finds $q_i$ and $p_u$ that minimize the following objective, where $\lambda$ is the regularization parameter and $\mathbb{T}$ is the training set:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2\right).
\end{equation*}

It is also common to include user biases $b_u$ and item biases $b_i$ in 
the optimization, defined as the average deviation of user $u$ and item 
$i$ to the global average $\mu$. The problem is then to find $q_i$ and $p_u$ that minimize the 
following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - \mu - b_u - b_i - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2 + b_u^2 + b_i^2\right)
\end{equation*}

Stochastic gradient descent and alternating least squares (ALS) are 
often used as optimization techniques. In the remainder, we use ALS as 
implemented in Apache Spark's MLLib version 2.3.0. We also round predictions 
to the nearest integer to obtain binary values.

@@

\begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
                  \caption{Complete Columns}
                  \label{fig:Columns-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
                  \caption{Complete Rows}
                  \label{fig:Rows-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                 \includegraphics[width=\columnwidth]{figures/5vs7_RS_04_training}
                \caption{Random Subjects (RS)}
                  \label{fig:Uniform-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_RFNU_04_training}
                  \caption{RFNU}
                  \label{fig:Diagonal-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_RFNT-L_04_training}
                  \caption{RFNTL}
                  \label{fig:triangular-L-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_RFNT-S_04_training}
                  \caption{RFNTS}
                  \label{fig:triangular-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.45\columnwidth]{figures/5vs7_random-unreal_04_training}
                  \caption{Random Unreal}
                  \label{fig:Random-Unreal-Sample-Training-set}
        \end{subfigure}
        \caption{Training sets for different sampling methods; $\alpha=0.4$ (dark elements are in the training set).}
        \label{fig:sampling}
\end{figure}

@@

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]
 For $\alpha \leq 0.5$, we ensure that the average number of selected 
 files by subject is $\alpha N_f$, by sampling the number of 
 files in every subject from $U(0,N_f)$ with 
 probability $2\alpha$, and setting it to 0 otherwise. For $\alpha > 
 0.5$, this is ensured by the value of $a$, which leads the 
 average of U(a, b) to be $\alpha N_f$.

@@

\subsubsection{Random Unreal (Figure 
~\ref{fig:Random-Unreal-Sample-Training-set})} The training set is 
sampled in a random uniform way, regardless of the file generation 
times. This method does not respect the time constraints in 
Equation~\ref{eq:time}: it will be used as a baseline for comparison 
with other methods.

@@

We generated synthetic matrices as shown in 
Figure~\ref{fig:synthetic-data}. Each matrix has 100 files and 100 
subjects of different \emph{types}. Subjects of the same type behave 
identically and all types contain the same number of subjects $\pm$ 1. 
Such a decomposition by subject type correspond to the case where 
different subjects may have data of different nature, as is the case in the
dataset of the Human Connectome Project~\cite{van2013wu} where not all the 
subjects have the same amount of images.

 For subjects of a given type, the matrix consists of 
 $log(n)$  blocks, where $n$ is the number of types. Blocks are 
 defined with all the possible variation patterns: some types do not 
 vary at all, while other ones vary in every block. Such 
 patterns are meant to mimic the logic of data 
processing pipelines. Each block of files represents the files produced 
at a given stage of the pipeline, which may or may not contain 
reproducibility errors depending on the subject type.

@@

Figure~\ref{fig:utility-matrices} shows the matrices
for PreFreesurfer and Freesufer. The
reproducibility of these pipelines varies across subjects,
but most files behave
consistently across all subjects, leading to complete black or white
lines.
%The ratio of positive elements in utility matrix of CentOS5 vs CentOS6(C5C6),  CentOS5 vs CentOS7(C5C7) and  CentOS6 vs CentOS7(C6C7) are
%$0.34$, $0.79$ and $0.79$ respectively.

@@

\section{Results}
\label{sec:results}
We conducted two experiments for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with subject and file biases. We 
compare the performance of our sampling methods to (1) a dummy 
classifier that always predicts the value in the majority class and (2) 
the Random Unreal method, used as the baseline sampling technique. All 
reported values are averages over 5 repetitions. Due to sampling 
issues, it is possible that the actual training ratio obtained with 
some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true. The code and data used to obtain the results are available through GitHub at
\url{https://github.com/big-data-lab-team/paper-reproducibility-collaborative-filtering}.
\todo{Soudabeh: push your code to reprotools. Tristan: release and check.}
\todo{publish code on zenodo}

@@

\subsubsection{ALS without Bias}
 Figure~\ref{fig:results-synthetic} shows accuracy results for ALS 
 without bias. By construction, the accuracy of the dummy classifier is 
 close to 0.5 for all subject types. Random Unreal performs well 
 for all subject types, which confirms that ALS is working correctly. 
 All the other methods 
perform better than the dummy classifier, and their accuracy decreases 
as the number of subject types increases.
However, only 3 methods provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. 
Surprisingly, RFNTS does not perform well for more than 2 subject 
types, even for $\alpha=0.9$.

@@

Figure~\ref{fig:results-synthetic-als-bias} shows accuracy results for 
ALS with bias. RS, RFNU and RFNTL remain the methods that best compare 
to Random Unreal, but their accuracy is much lower than without 
biases. This is presumably due to 
the fact that file biases are all close to 0.5. 
In such a situation, biases are detrimental and should not be included.

@@

Figure~\ref{fig:results-real-als} shows the accuracy for ALS without bias. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets when the training ratio is larger 
than 0.5. Complete rows and complete columns do not reach the accuracy of 
the dummy classifier. Random Unreal has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.

@@

Figure~\ref{fig:results-real-als-bias} shows the accuracy for ALS with 
bias. From a training ratio of 0.5, all methods perform well for all 
datasets, except complete columns in 
Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, RFNTL and 
RFNTS also perform well for a training ratio lower than 0.5. Overall, 
the accuracy is much higher than without bias, due to the fact that the 
file biases are very strong. In the remainder, all the 
results are obtained using ALS without bias.

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{data/results/roc/roc.pdf}
\includegraphics[width=\columnwidth]{data/results/roc/roc-closeup.pdf}
\caption{Comparison of the sampling methods in the ROC space for the 6 synthetic datasets and Freesurfer (ALS without bias). Right: entire dataset. Left: close-up on the top-left part.}
\label{fig:roc}
\end{figure*}

\begin{table}
\centering
\begin{tabular}{ccc}
& Sensitivity & Specificity \\
\hline
Complete Columns & 0.53 & 0.97\\
Complete Rows & 0.43 & 0.89\\
Random Subjects & 0.88 & 0.99\\
RFNU & 0.92 & 0.99 \\
RFNTL & 0.81 & 0.97\\
RFNTS & 0.65 & 0.93 \\
Random Unreal & 1 & 1
\end{tabular}
\caption{Average sensitivity and specificity for the synthetic datasets and Freesurfer ($\alpha=0.9$, ALS without bias).}
\label{table:roc}
\end{table}

\subsection{Effect of the number of factors} Figure~\ref{fig:factors} 
shows the effect of the number of factors used in the ALS optimization 
for the RFNU method and the synthetic dataset with 8 subject types (3 
blocks of files). For training ratios greater than 0.5, the accuracy 
with 3 factors is substantially higher than with 2 factors. Beyond 3, 
the number of factors does not have any effect, which shows that the 
data is best 
explained by 3 factors. Figure~\ref{fig:factors3d} confirms that 3 
factors are enough to separate the files of this dataset in 3 
blocks, and the subjects in 8 types.

\subsection{Effect of the maximum number of iterations}
Figure~\ref{fig:iterations} illustrates the effect of the number of 
iterations used in the ALS optimization, for the 
RFNU method and the synthetic dataset with 8 types. 
From a training ratio of 0.7, the number of iterations has a moderate 
effect on the accuracy. We used 5 iterations in our experiments, 
which slightly degrades the accuracy compared to 15 or 20.

@@

Figure~\ref{fig:overlap} compares the prediction errors made by RFNU, 
RFNTL and RFNTS. The training set is represented
in black (negatives) and white (positives), and the test set is in green 
(true positives), yellow (false negatives), gray (true negatives) and 
red (false positives). This representation provides insights regarding 
where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of prediction errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

Figure~\ref{fig:error-locality} shows the results obtained by RFNU on 
the Freesurfer dataset for a training ratio of 0.6. The prediction 
error is localized (1) at the bottom of the matrix, which corresponds 
to the end of the execution, and (2) in the regions where file 
reproducibility varies across subjects, i.e., lines are not entirely 
black or entirely white.
